---
title: "Carseats Data Analysis"
author: "Stefany Ayala"
output: github_document
---

```{r setup, include=FALSE}
# Set up the environment and options
knitr::opts_chunk$set(echo = TRUE)
```

## Data Preparation

```{r}
# Define the URL of the Rdata file
url <- "https://drive.google.com/uc?id=16-zxtGm7-qYcjEKuxbLpZ3-nLYW-WAzX&export=download"

# Define a temporary file path to store the downloaded Rdata file
temp_file <- tempfile(fileext = ".Rdata")

# Download the Rdata file
download.file(url, destfile = temp_file, mode = "wb")  # Use mode="wb" for binary files

# Load the Rdata file into the environment
load(temp_file)

# Assign the dataset to a variable
df <- Data

# Display the first few rows of the dataset
head(df)
```

## Model Building and Evaluation

```{r, message=FALSE, warning=FALSE}
# Load required libraries
library(randomForest)
library(klaR)
library(caret)
library(ROCR)
library(e1071)
library(rpart)
library(MASS)
library(KernelKnn)
library(FNN)
```

### Part I: Gain Chart

#### 1. Build Random Forest and Naïve Bayes models to predict the variable `Target`

```{r}
set.seed(123)  # Set seed for reproducibility

# Split the data into training and testing sets (80% training, 20% testing)
index <- sample(2, nrow(df), replace = TRUE, prob = c(0.8, 0.2))
train_data <- df[index == 1,]
test_data <- df[index == 2,]

# Random Forest Model
rf <- randomForest(Target ~ ., data = train_data, ntree = 300, proximity = TRUE,
                   replace = TRUE, importance = TRUE)

# Naive Bayes Model
nb <- naiveBayes(Target ~ ., data = train_data)
nb_score <- predict(nb, test_data, type = 'raw')
```

#### 2. Gain Chart for `Target = Yes`

```{r}
# Random Forest Prediction
rff <- predict(rf, test_data, type = 'vote')
score <- rff[, 2]

# Naive Bayes Prediction
pred_rf <- prediction(score, test_data$Target)
perf_rf <- performance(pred_rf, "tpr", "rpp")

pred_nb <- prediction(nb_score[, 2], test_data$Target)
perf_nb <- performance(pred_nb, "tpr", "fpr")

# Extract gain values for Naive Bayes
gain.x.nb <- unlist(slot(perf_nb, 'x.values'))
gain.y.nb <- unlist(slot(perf_nb, 'y.values'))

# Plot Gain Chart
plot(perf_rf, col = 'red', lty = 1, lwd = 2, main = "Gain Chart for RF and Naive Bayes Models")
lines(x = gain.x.nb, y = gain.y.nb, col = "orange", lwd = 2)
legend("bottomright", legend = c("Random Forest", "Naive Bayes"), lty = c(1, 1),
       col = c("red", "orange"), bty = "n", cex = 0.75)
```

**Conclusion**: The Naive Bayes model is preferable based on the higher area under the curve indicating better model performance.

#### 3. Gain Chart for `Target = No`

```{r}
# Random Forest Prediction for 'Target = No'
rff <- predict(rf, test_data, type = 'vote')
score <- rff[, 1]

# Naive Bayes Prediction for 'Target = No'
pred_rf <- prediction(score, test_data$Target)
perf_rf <- performance(pred_rf, "tpr", "rpp")

pred_nb <- prediction(nb_score[, 1], test_data$Target)
perf_nb <- performance(pred_nb, "tpr", "fpr")

# Extract gain values for Naive Bayes
gain.x.nb <- unlist(slot(perf_nb, 'x.values'))
gain.y.nb <- unlist(slot(perf_nb, 'y.values'))

# Plot Gain Chart for 'Target = No'
plot(perf_rf, col = 'red', lty = 1, lwd = 2, main = "Gain Chart for RF and Naive Bayes Models")
lines(x = gain.x.nb, y = gain.y.nb, col = "orange", lwd = 2)
legend("bottomright", legend = c("Random Forest", "Naive Bayes"), lty = c(1, 1),
       col = c("red", "orange"), bty = "n", cex = 0.75)
```

#### 4. Build a Decision Tree Model

```{r}
# Decision Tree Model
dtree <- rpart(Target ~ ., data = train_data, method = "class")
```

#### 5. Model Evaluation using Confusion Matrix

```{r}
# Evaluate Random Forest Model
confusionMatrix(predict(rf, type = "class", newdata = test_data),
                test_data$Target, positive = "YES")

# Evaluate Naive Bayes Model
confusionMatrix(predict(nb, type = "class", newdata = test_data),
                test_data$Target, positive = "YES")

# Evaluate Decision Tree Model
confusionMatrix(predict(dtree, type = "class", newdata = test_data),
                test_data$Target, positive = "YES")
```

**Conclusion**: Based on accuracy, both the Random Forest and Naive Bayes models are preferable, as they have similar performance.

#### 6. Model Preference Based on Cost of Misclassification

In the case where misclassifying stores with low sales has a higher cost than misclassifying stores with high sales, the **Random Forest model** is preferred. This is because Random Forest is better at minimizing false negatives, which can help reduce the cost of misclassifying low-sales stores as high-sales ones.

#### 7. Model Preference for Classifying High Sales Stores

If the priority is correctly classifying stores with high sales, the **Naive Bayes model** is preferable, as it performs better in terms of precision for classifying "high" sales stores.

---

### Part II: K-Nearest Neighbors (KNN)

#### 1. Prepare the Dataset for KNN

```{r}
# Min-max normalization function
normalize <- function(x) {
  num <- x - min(x)
  denom <- max(x) - min(x)
  return (num / denom)
}

# Apply normalization to the first 5 columns of the dataset
df_norm <- as.data.frame(lapply(df[1:5], normalize))
```

#### 2. Build a KNN Model with k = 3

```{r}
# Split normalized data into training and test sets
df.training <- df_norm[index == 1, 1:5]
df.test <- df_norm[index == 2, 1:5]
df.trainLabels <- df[index == 1, 11]
df.testLabels <- df[index == 2, 11]

# Build KNN model with k=3
df_pred_k3 <- knn(train = df.training, test = df.test, cl = df.trainLabels, k = 3, prob = TRUE)
```

#### 3. Build a KNN Model with k = 1

```{r}
# Build KNN model with k=1
df_pred_k1 <- knn(train = df.training, test = df.test, cl = df.trainLabels, k = 1, prob = TRUE)
```

#### 4. Evaluate the Models Based on Accuracy

```{r}
# Train KNN model using caret package for k values 1 to 3
model <- train(Target ~ ., data = df, method = 'knn', tuneGrid = expand.grid(k = 1:3), metric = 'Accuracy')

# Display the model results
model
```

**Conclusion**: Based on accuracy, it is better to use **k=3** as the optimal number of neighbors for the KNN model.

```
