---
title: "Pima Indian Diabetes Analysis"
author: "Stefany Ayala"
output: github_document
---

# Introduction
The objective of this analysis is to explore, clean, and prepare the **Pima Indian Diabetes dataset** for modeling. This report demonstrates the application of data cleaning, visualization, and machine learning techniques in R. Additionally, we compare the performance of two algorithms to showcase modeling versatility.

---

# 1. Data Loading
We first load the data directly from a URL into R and inspect its structure.

```{r, message=FALSE}
# Load necessary libraries
library(ggplot2)
library(OneR)
library(rpart)

# Load the dataset
url <- "https://drive.google.com/uc?id=1MvxSOTYHknWFIukVGaEzR79-cLX_JHMC&export=download"
diabetes <- read.csv(url)

# Inspect dataset structure
dim(diabetes)  # Dimensions of the dataset
attributes(diabetes)  # Attributes of the dataset
```
# 2. Data Exploration
## 2.1 Column Renaming
For better readability, we rename the columns.

```{r, message=FALSE}
# Rename columns
colnames(diabetes) <- c('pregnant', 'glucose', 'bp', 'triceps', 'insulin', 'bmi',
                        'diabetes', 'age', 'target')
```
## 2.2 Initial Data Insights
We explore the dataset's variables with histograms and basic descriptive statistics.

```{r, message=FALSE}
# Visualize data distributions
par(mfrow = c(3, 3))
for (col in names(diabetes)[-9]) {
  hist(diabetes[[col]], main = paste("Histogram of", toupper(col)), xlab = toupper(col))
}

# Summarize descriptive statistics
summary(diabetes)
```

# 3. Data Cleaning
## 3.1 Detecting and Handling Outliers
Outliers are detected using Tukey’s method and replaced with NA.

```{r, message=FALSE}
# Handle outliers
clean_data <- diabetes
for (col in names(diabetes)[-9]) {
  outliers <- boxplot.stats(diabetes[[col]])$out
  clean_data[[col]] <- ifelse(diabetes[[col]] %in% outliers, NA, diabetes[[col]])
}
```
## 3.2 Imputing Missing Values
Missing values are replaced with the mean of each variable.

```{r, message=FALSE}
# Impute missing values with the mean
clean_data <- as.data.frame(lapply(clean_data, function(x)
  ifelse(is.na(x), mean(x, na.rm = TRUE), x)))

# Verify no missing values remain
sum(is.na(clean_data))
```

# 4. Data Preparation
The cleaned dataset is split into training (80%) and testing (20%) sets.
```{r, message=FALSE}
# Split data
set.seed(123)
train_indices <- sample(1:nrow(clean_data), size = 0.8 * nrow(clean_data))
TrainData <- clean_data[train_indices, ]
TestData <- clean_data[-train_indices, ]
```

# 5. Modeling
Interpretation

- **OneR Model:**

This simple model provides interpretable rules based on a single attribute, which can be ideal for quick insights.
The model's performance metrics indicate its effectiveness in this context.

- **Decision Tree Model:**

The confusion matrix reveals the strengths and weaknesses in classification.
Decision Trees, though more complex, offer better granularity in predictions.

By comparing the results of both models, we gain insights into the trade-offs between model simplicity and performance.

## 5.1 OneR Model
We train a OneR model to identify the best attribute and evaluate its performance.
```{r, message=FALSE}
# Train a OneR model
one_r_model <- OneR(target ~ ., data = TrainData, verbose = TRUE)

# Summarize the model
summary(one_r_model)

# Predict on test data
one_r_predictions <- predict(one_r_model, TestData)

# Evaluate the model
one_r_eval <- eval_model(one_r_predictions, TestData)
one_r_eval
```

## 5.2 Decision Tree Model
As a comparison, we train a Decision Tree model and evaluate its performance.
```{r, message=FALSE}
# Train a decision tree model
tree_model <- rpart(target ~ ., data = TrainData, method = "class")

# Predict on test data (specifying 'type = "class"' to get class labels)
tree_predictions <- predict(tree_model, TestData, type = "class")

# Confusion matrix
confusion_matrix <- table(Predicted = tree_predictions, Actual = TestData$target)
confusion_matrix
```

# 6. Results and Evaluation
## 6.1 OneR Model Results
The OneR model was trained on the dataset to identify the best single predictor of the target variable. The OneR model achieves an accuracy of **75.24%** based on the rule for the `glucose` variable. The model divides the `glucose` variable into ranges and predicts the target variable based on these ranges.

The confusion matrix is as follows:

```{r, message=FALSE}

# Display evaluation metrics
one_r_eval
```
From the confusion matrix, we can see that the OneR model has some false positives and false negatives, but it is still fairly accurate for this task.

## 6.2 Decision Tree Model Results
To compare the performance of a more complex algorithm, a Decision Tree model was trained. Its predictions on the test data are evaluated using a confusion matrix:

```{r, message=FALSE}
# Confusion matrix for Decision Tree model
confusion_matrix <- table(tree_predictions, TestData$target)

# Display confusion matrix
confusion_matrix
```
The Decision Tree model is capable of capturing more complex relationships in the data compared to the OneR model, though it also has some misclassifications.

# Final thoughs
Both models show reasonable accuracy, but the OneR model is simpler and more interpretable. The Decision Tree model, while more complex, may be more suitable for capturing non-linear patterns. Further experimentation, such as cross-validation and hyperparameter tuning, could lead to improved performance.
